Exploring probability distributions for bivariate temporal granularities

### 2019-11-03 main.Rmd
cleaning old ggplot2 codes to prob_plot

<!--temporal granularities and why should we care --> 
Temporal data can be available at various resolution depending on the context. Social and economic data are often collected and reported at coarser temporal scales like monthly, quarterly or annually. But with recent advancement in technology, more and more data are recorded and stored at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be collected at every second. As the frequency of data increases, the number of questions about the observed variable that need to be addressed also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways.

A temporal granularity which results from such a deconstruction may be intuitively described as a sequence of time granules, each one consisting of a set of time instants. There are several classes of time deconstructions including linear time granularities, circular time granularities and aperiodic calendar categorizations. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Circular time granularities accommodate periodicities in time such as hour of the day, and day of the week. Aperiodic calendar categorizations are neither linear nor circular, such as day of the month or public holidays.

It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing.

@@

<!--details abt time gran --> 
The hierarchical structure of many granularities creates a natural nested ordering. For example, hours are nested within days, days within weeks, weeks within months, and so on. We refer to granularities which are nested within multiple levels as "multiple-order-up" granularities. For example, hour of the week and second of the hour are both multiple-order-up, while hour of the day and second of the minute are single-order-up.

@@

@@

Even with harmonies, problems might occur due to rarely occuring categories. suppose we have $T$ hourly observations, and $C_1$ is HOD and $C_2$ is DOW. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For scatterplots, even $k=3$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

@@

```{r moy-dom,echo=FALSE, out.width = "70%", fig.align = 'center',fig.cap ="Letter-value plot of electricity consumption in Victoria across days of the month\n faceted by months January, February, April and December. "}

# change all of these to gravitas function
VIC <- read.csv("data/VIC2015/Data/demand_VIC.csv")
VIC$Date <- as.Date(VIC$Date, origin = "1899-12-30")

first_day_of_month_wday <- function(dx) {
  day(dx) <- 1
  wday(dx)
}

VIC <- VIC %>% mutate(
  Indx_Year = year(Date),
  Indx_Month = month(Date, label = FALSE, abbr = TRUE),
  Indx_Wk_Yr = week(Date),
  Indx_Wk_Month = ceiling((day(Date) + first_day_of_month_wday(Date) - 1) / 7),
  Indx_Day_Week = wday(Date,
    label = FALSE, abbr = TRUE,
    week_start = 1
  ),

  Indx_Day_Month = day(Date),

  Indx_Day_Year = yday(Date),

  Indx_Weekend = if_else(Indx_Day_Week %in% c(6, 7), 1, 0),

  Indx_HlHr_day = Period,
  month = month(Date, label = FALSE, abbr = TRUE),
  year = year(Date),
  yday = yday(Date),
  wday = wday(Date,
    label = FALSE, abbr = TRUE,
    week_start = 1
  ),
  bow = (wday - 1) * 48 + Period,
  dom = day(Date),
  bom = (dom - 1) * 48 + Period,
  Weekend = if_else(wday %in% c(6, 7), 1, 0),
  Indx_hour = ceiling(Period / 2),
  Indx_Hour_Yr = Indx_hour + 24 * (yday - 1),
  Indx_Hour_Month = Indx_hour + 24 * (Indx_Day_Month - 1),
  Indx_Hour_Wk = Indx_hour + 24 * (wday - 1)
)

VIC <- as_tibble(VIC)
```

```{r allFig, fig.height=5.5, out.width="50%", fig.pos = "p", fig.align= 'left',echo=FALSE, eval=TRUE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by DoM and MoY, (b) Decile plot by HoD and DoW (c) Box plot by DoY and DoM, (d) Violin plot of DoM and WoM, (e) Ridge plot by WoM and WoY, (f) Decile plot by DoY and MoY. Only plots (a) and (b) show harmonised time variables.", fig.show = 'hold'}

VIC %>% filter(year %in% c(2012, 2013, 2014), Indx_Month %in% c(1, 2, 4, 12)) %>% ggplot(aes(Indx_Day_Month, OperationalLessIndustrial, group = Indx_Day_Month)) + geom_lv(aes(fill = ..LV..), outlier.colour = "red", outlier.shape = 1) + scale_fill_brewer() + facet_wrap(~Indx_Month) + ylab("Electricity Demand [KWh]") + xlab("Days of the Month") + scale_x_continuous(breaks = seq(0, 31, 5)) + ggtitle("(a) Letter value plot")

VIC_hod_dow <- VIC %>%
  filter(year %in% c(2012, 2013, 2014)) %>%
  group_by(wday, Indx_hour) %>%
  do({
    x <- .$OperationalLessIndustrial
    map_dfr(
      .x = seq(0.1, 0.9, 0.1),
      .f = ~ tibble(
        Quantile = .x,
        Value = quantile(x, probs = .x, na.rm = TRUE)
      )
    )
  }) %>%
  filter(wday %in% c(1, 2, 6, 7))

VIC_hod_dow %>% ggplot(aes(x = Indx_hour, y = Value, col = as.factor(Quantile))) + geom_line() + facet_wrap(~wday) + scale_x_continuous(breaks = seq(1, 24, 5)) + ylab("") + xlab("Day of the Week") + theme(legend.position = "none", strip.text = element_text(size = 7, margin = margin())) + ggtitle("(b) Decile plot")

VIC %>%
  filter(year %in% c(2012, 2013, 2014), Indx_Day_Month %in% c(1, 15, 29, 31)) %>%
  ggplot(aes(as.factor(yday), OperationalLessIndustrial, group = yday)) + geom_boxplot() + facet_wrap(~Indx_Day_Month) + ylab("Electricity Demand [KWh]") +
  xlab("Days of the Year") + scale_x_discrete(breaks = seq(0, 366, 60)) + theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + ggtitle("(c) Box plot")

VIC %>% filter(year %in% c(2012, 2013, 2014), Indx_Wk_Month %in% c(1, 2, 4)) %>% ggplot(aes(as.factor(Indx_Day_Month), OperationalLessIndustrial)) + geom_violin(alpha = 0.03) + facet_wrap(~Indx_Wk_Month, nrow = 3) + ylab("") + xlab("Days of the Month") + theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + scale_x_discrete(breaks = seq(0, 31, 5)) + scale_y_continuous(breaks = seq(2000, 9000, 2000)) + ggtitle("(d) Violin plot")

VIC %>% dplyr:::filter(year %in% c(2012, 2013, 2014), Indx_Wk_Month %in% c(1, 2, 5), Indx_Wk_Yr < 20) %>% ggplot(aes(x = OperationalLessIndustrial, y = as.factor(Indx_Wk_Yr), group = Indx_Wk_Yr)) + geom_density_ridges2() + facet_wrap(~Indx_Wk_Month) + xlab("Electricity Demand [KWh]") + ylab("Weeks of the Year") + scale_x_continuous(breaks = seq(2000, 10000, 3000)) + theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + ggtitle("(e) Ridge plot")

VIC_moy_doy <- VIC %>%
  filter(year %in% c(2012, 2013, 2014)) %>%
  group_by(Indx_Month, yday) %>%
  do({
    x <- .$OperationalLessIndustrial
    map_dfr(
      .x = seq(0.1, 0.9, 0.1),
      .f = ~ tibble(
        Quantile = .x,
        Value = quantile(x, probs = .x, na.rm = TRUE)
      )
    )
  }) %>%
  filter(Indx_Month %in% c(1, 7, 11))

VIC_moy_doy %>% ggplot(aes(x = yday, y = Value, col = as.factor(Quantile), group = yday)) + geom_line() + facet_wrap(~Indx_Month) + scale_x_continuous(breaks = seq(1, 336, 60)) + ylab("") + xlab("Day of the Year") + theme(legend.position = "none", strip.text = element_text(size = 7, margin = margin())) + ggtitle("(f) Decile plot")

# grid.arrange(p1,p2,p3,p4,nrow = 2)

>>>>

<!--temporal granularities and why should we care -->

<!-- problem of analyzing data with such finer scales and classical ways to deal with it -->
Temporal data are available at various resolution depending on the context. Social and economic data like GDP are often collected and reported at coarser temporal scales like monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded and stored at much finer temporal scales. Energy consumption is collected every half an hour, while energy supply is collected every minute and web search data might be collected at every second. As the frequency of data increases, the number of questions about periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales like days, months or quarters. This approach requires deconstructing time in various possible ways. Calendar-based graphics[] can unpack the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. They mainly pick out patterns in the weekly and monthly structure well and are capable of checking the weekends or special days.  Any sub-daily resolution temporal data can also be displayed using this type of faceting [reference trellis plot] with days of the week, month of the year and another sub-daily deconstruction of time.

<!-- # challenges of classical ways of displaying -->
But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. A temporal granularity which results from such a deconstruction may be intuitively described as a sequence of time granules, each one consisting of a set of time instants[]. There can be several classes of time deconstructions including linear time granularities, circular time granularities and aperiodic calendar categorizations. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Circular time granularities accommodate periodicities in time such as hour of the day, and day of the week. Aperiodic calendar categorizations are neither linear nor circular, such as day of the month or public holidays. Also, <!--details abt time gran --> 
the hierarchical structure of time creates a natural nested ordering. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, we can construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour.

<!-- We refer to granularities which are nested within multiple levels as "multiple-order-up" granularities. For example, hour of the week and second of the hour are both multiple-order-up, while hour of the day and are single-order-up.  -->

<!-- TODO -->
It is important to be able to navigate through all of these temporal granularities to have multiple perspectives on the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing.

@@

@@

```{r allFig, fig.height=5.5, out.width="50%", fig.pos = "p", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by DoM and MoY, (b) Decile plot by HoD and DoW (c) Box plot by DoY and DoM, (d) Violin plot of DoM and WoM, (e) Ridge plot by WoM and WoY, (f) Decile plot by DoY and MoY. Only plots (a) and (b) show harmonised time variables.", fig.show = 'hold'}

# change all of these to gravitas function
library(gravitas)
library(tsibbledata)
library(dplyr)
library(ggplot2)

VIC <- vic_elec

VIC %>% 
  create_gran("month_year") %>% 
  filter(month_year %in% c("Jan", "Mar", "July", "Dec")) %>% 
  prob_plot("month_year","day_month",
            response = "Demand",
            plot_type = "lv") + 
  ylab("Electricity Demand [KWh]") +
  ggtitle("(a) Letter value plot")

VIC %>% 
  create_gran("day_week") %>% 
  filter(day_week %in% c("Mon", "Tue", "Sat", "Sun")) %>% 
  prob_plot("day_week","hour_day",
            response = "Demand",
            plot_type = "quantile",
            symmetric = FALSE) + 
  ylab("Electricity Demand [KWh]") +
  ggtitle("(b) Decile plot")

VIC %>%
  create_gran("day_month") %>% 
  filter(day_month %in% c(1, 15, 29, 31)) %>%
  prob_plot("day_month",
                "day_year",
                response = "Demand",
                plot_type = "boxplot")+
  ylab("Electricity Demand [KWh]") +
  scale_x_discrete(breaks = seq(0, 366, 60)) + 
  theme(legend.position = "bottom", 
        strip.text = element_text(size = 7, margin = margin())) +
  ggtitle("(c) Box plot")

VIC %>%
    create_gran("week_month") %>% 
  filter(week_month %in% c(1, 2, 4)) %>%
   prob_plot("week_month",
                "day_month",
                response = "Demand",
                plot_type = "violin", alpha = 0.3) + ylab("") + 
  xlab("Days of the Month") + 
  theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + scale_x_discrete(breaks = seq(0, 31, 5)) + 
  scale_y_continuous(breaks = seq(2000, 9000, 2000)) +
  ggtitle("(d) Violin plot")

VIC %>%  create_gran("week_month") %>%
  create_gran("week_year") %>%
  filter(week_month%in% c(1, 2, 5), week_year %in% c(1:20)) %>% 
  prob_plot("week_year", 
            "week_month",
            response = "Demand",
            plot_type = "ridge")  + 
  xlab("Electricity Demand [KWh]") +
  ylab("Weeks of the Year") +
  scale_x_continuous(breaks = seq(2000, 10000, 3000)) + 
  theme(legend.position = "bottom", 
        strip.text = element_text(size = 7, margin = margin())) +
  ggtitle("(e) Ridge plot")

VIC %>%
  create_gran("month_year") %>% 
  filter(month_year %in% c("Jan", "Jul", "Nov")) %>% 
   prob_plot("month_year", 
            "day_year",
            response = "Demand",
            plot_type = "quantile",
            quantile_prob = seq(0.1, 0.9, 0.1),
            symmetric = FALSE) +
   #scale_x_discrete(breaks = seq(1, 336, 60)) + 
   ylab("") + 
   xlab("Day of the Year") + theme(legend.position = "none", 
                                   strip.text = element_text(size = 7, 
                                                             margin = margin())) +
  ggtitle("(f) Decile plot")

# VIC_moy_doy %>% ggplot(aes(x = yday, y = Value, col = as.factor(Quantile), group = yday)) + geom_line() + facet_wrap(~Indx_Month) + scale_x_continuous(breaks = seq(1, 336, 60)) + ylab("") + xlab("Day of the Year") + theme(legend.position = "none", strip.text = element_text(size = 7, margin = margin())) + ggtitle("(f) Decile plot")
```

@@

Even with harmonies, problems might occur due to rarely occuring categories.

suppose we have $T$ hourly observations, and $C_1$ is HOD and $C_2$ is DOW. Each element of $C_1$ occurs approximately $T/n_1$ times while each element of $C_2$ occurs approximately $T/n_2$ times. There are no structurally empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(n_1n_2)$. If we require at least $k$ observations to create a meaningful panel, then provided $T \ge n_1n_2k$, the visualization will be acceptable. The value of $k$ will depend on what type of visualization we are producing. For scatterplots, even $k=3$ may be acceptable, but for density estimates, we probably need $k \ge 30$.

@@

```{r}
# grid.arrange(p1,p2,p3,p4,nrow = 2)
